Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-22 analysis/paper/paper.Rmd
reduce code in paper

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      #before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    #extra_dependencies: "subfig"
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

@@

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
were extrapolated (see Supplementary material for more details). For potato
treatment samples, the whole slide was counted.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. Wheat and mix counts were extrapolated as described above.
Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

@@

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg 
calculus (also z-score standardised) to account for differences in starch 
concentration in the calculus [as per 
@wesolowskiEvaluatingMicrofossilContent2010].

@@

```{r solution-count-out}
count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

### adding the 'both' row to the data frame ###
count_sol_comb_out <- count_sol_comb %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
  ungroup() %>%
  add_row(tibble_row(count_sol_comb_out), .before = 3)

count_sol_out[1,3] <- NA
```

@@

```{r sample-count}
count_sample <- count_sample %>%
  ungroup() %>%
  add_row(count_sample_comb, .after = 2)

# convert columns to contain mean count and percentage of total count
count_samp_per <- count_sample %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))

count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
  select(1,2,3,7,4,8,5,9,6,10) #%>%

count_samp_out[1,3] <- NA # replace with NA
  #add_row(tibble_row(count_sample_comb_out), .before = 3)
```

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this 
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the total amount of small starches in these two solutions, 
`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
are from wheat.
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

```{r solution-count-tbl}
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
options(knitr.kable.NA = "")
knitr::kable(count_sol_out, col.names = cnames, 
             caption = "Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.")
```

```{r sample-count-tbl}
cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
            "Large (%)", "SD", "Total (%)", "SD")
knitr::kable(count_samp_out, col.names = cnames, 
             caption = "Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.")
```

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(subset(count_sol, solution == "mix")$total, 3)`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r signif(subset(count_sample, treatment == "wheat")$total, 3)`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r signif(subset(count_sol, solution == "potato")$total, 3)`)
and in the biofilm samples 
(`r signif(subset(count_sample, treatment == "potato")$total, 3)`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

@@

```{r sample-perc}
# convert proportions to percent labels
perc_comb <- prop_comb %>% 
  mutate(across(c(s, m, l, total), scales::percent, 0.001))

perc <- prop %>%
  mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
  add_row(tibble_row(perc_comb), .before = 3)
```

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorporated into the samples.")
```

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc$total), max(perc$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e., potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

<!-- mixed treatments -->
Wheat incorporation was most affected in the mixed-treatment samples, with only
`r paste0(perc$total[2])` of the total available starches being incorporated into
the sample, compared to
`r paste0(perc$total[4])`
in the separated wheat treatment.

@@

Overall, medium starch granules had a higher mean rate of incorporation 
(`r scales::percent(mean(prop$m, na.rm = T), 0.001)`) 
than small 
(`r scales::percent(mean(prop$s, na.rm = T), 0.001)`) 
and large 
(`r scales::percent(mean(prop$l, na.rm = T), 0.001)`) 
starch granules across all treatments, while large potato starches had the lowest 
rate of incorporation across all treatments.

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(diff_sep$potato[3]), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(diff_mix$potato[3]), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc$total), max(perc$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(count_sol$total[4], big.mark = ",")`) 
and potato
(mean = `r format(count_sol$total[3], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc$total), max(perc$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], <!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies -->which could in part be due to the 
lack of oral amylase activity in our model.

@@

<!-- If you're reading this, thank you for taking the time to read the source 
documents. You're awesome! -->
We would like to thank Dr. Stephanie Schnorr for help with the amylase activity protocol.
We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
we will NOT acknowledge the use of Sci-Hub in this study.

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      #before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    #extra_dependencies: "subfig"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

@@

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
were extrapolated to the total number of granules exposed to the samples over 16
days of treatments (see Supplementary Material for more details). For potato
treatment samples, the whole slide was counted.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material).

Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

@@

Pearson's *r* was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg 
calculus (also z-score standardised) to account for differences in starch 
concentration in the calculus [as per 
@wesolowskiEvaluatingMicrofossilContent2010].

@@

```{r solution-count-out}
# sol_comb_row <- sol_comb_out %>%
#   group_by(treatment) %>%
#   filter(treatment == "mix") %>%
#   summarise(starch = "both",
#             across(where(is.numeric), sum, na.rm = T))

sol_tbl <- sol_comb_out %>%
  #ungroup() %>%
  #add_row(sol_comb_row, .after = 2) %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

# count_sol_out <- count_sol %>%
#   mutate(across(c(s, m, l, total), 
#                 function(x) 
#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
# 
# ### adding the 'both' row to the data frame ###
# count_sol_comb_out <- count_sol_comb %>%
#   mutate(across(c(s, m, l, total), 
#                 function(x) 
#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))
# 
# count_sol_out <- count_sol %>%
#   mutate(across(c(s, m, l, total), 
#                 function(x) 
#                   paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
#   ungroup() %>%
#   add_row(tibble_row(count_sol_comb_out), .before = 3)

#count_sol_out[1,3] <- NA
sol_tbl[1,3] <- NA
```

@@

```{r sample-count}
samp_counts <- corr_comb_long %>%
  pivot_wider(names_from = size, values_from = count) %>%
  ungroup() %>%
  add_row(select(samp_comb_row, !sample), .after = 2)

samp_sd_wide <- corr_sd_long %>%
  select(!count) %>%
  pivot_wider(names_from = size, values_from = sd) %>%
  ungroup() %>%
  add_row(samp_sd_row, .after = 2) %>%
  mutate(across(where(is.numeric), round)) %>%
  rename(s_sd = s,
         m_sd = m,
         l_sd = l,
         total_sd = total)

samp_counts_per <- samp_counts %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
                         ")")))

# combine samp_counts_per with samp_sd_wide
samp_counts_tbl <- samp_counts_per %>%
  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)

# count_sample <- count_sample %>%
#   ungroup() %>%
#   add_row(samp_sd_row, .after = 2)

# convert columns to contain mean count and percentage of total count
# count_samp_per <- count_sample %>%
#   mutate(across(c(s, m, l, total), 
#                 function(x) 
#                   paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))

# count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
#   select(1,2,3,7,4,8,5,9,6,10) #%>%

samp_counts_tbl[1,3] <- NA
#count_samp_out[1,3] <- NA # replace with NA
  #add_row(tibble_row(count_sample_comb_out), .before = 3)
```

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this 
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the total amount of small starches in these two solutions, 
`r paste0(signif((sol_comb_out$s[4] / sum(sol_comb_out$s[3], sol_comb_out$s[4])) * 100, 3), "%")`
are from wheat.
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

```{r solution-count-tbl}
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
options(knitr.kable.NA = "")
knitr::kable(sol_tbl, col.names = cnames, 
             caption = "Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.")
```

```{r sample-count-tbl}
cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
            "Large (%)", "SD", "Total (%)", "SD")
knitr::kable(samp_counts_tbl, col.names = cnames, 
             caption = "Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.")
```

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
and in the biofilm samples 
(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

@@

```{r sample-perc}
# convert proportions to percent labels
# perc_comb <- prop_comb %>% 
#   mutate(across(c(s, m, l, total), scales::percent, 0.001))

# perc <- prop %>%
#   mutate(across(c(s, m, l, total), scales::percent, 0.001)) %>% 
#   add_row(tibble_row(perc_comb), .before = 3)

prop_wide <- prop %>%
  ungroup() %>%
  pivot_wider(id_cols = c(treatment, starch), 
              names_from = size, 
              values_from = prop) %>%
  add_row(prop_comb_row, .after = 2)

perc_tbl <- prop %>%
  ungroup() %>%
  select(!c(sample, solution, prop)) %>%
  pivot_wider(names_from = size, values_from = perc) %>%
  add_row(perc_comb_row, .after = 2)
# samp_comb_row <- corr_comb_long %>%
#   pivot_wider(names_from = size, values_from = count) %>%
#   group_by(treatment) %>%
#   filter(treatment == "mix") %>%
#   summarise(starch = "both",
#             across(where(is.numeric), sum, na.rm = T)) %>%
#   mutate(sample = "sample")

# perc_comb_row <- sol_comb_row %>%
#   mutate(sample = "solution") %>%
#   add_row(samp_comb_row) %>%
#   pivot_longer(cols = c(s, m, l, total), names_to = "size")
#   mutate()

#perc <- prop %>%
#  pivot_wider(id_cols = c(treatment, starch), names_from = size, values_from = perc)
```

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc_tbl, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorporated into the samples.")
```

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e., potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

<!-- mixed treatments -->
Wheat incorporation was most affected in the mixed-treatment samples, with only
`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
the sample, compared to
`r paste0(perc_tbl$total[4])`
in the separated wheat treatment.

@@

```{r}
sol_size_prop <- sol_comb_out %>%
  group_by(treatment, starch) %>%
  #select(!total) %>%
  mutate(across(where(is.numeric), function(x) x / total)) %>%
  mutate(type = "solution")

samp_size_prop <- samp_counts %>%
  group_by(treatment, starch) %>%
  #select(!total) %>%
  mutate(across(where(is.numeric), function(x) x / total)) %>%
  mutate(type = "sample")

size_diff <- sol_size_prop %>%
  full_join(samp_size_prop) %>%
  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
  pivot_wider(names_from = type, values_from = count) %>%
  mutate(diff = (sample - solution) * 100) %>%
  select(!c(solution, sample)) %>%
  pivot_wider(names_from = size, values_from = diff) %>%
  select(!total)
```

Overall, medium starch granules had a higher mean rate of incorporation 
(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
than small 
(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
and large 
(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
starch granules across all treatments, while large potato starches had the lowest 
rate of incorporation across all treatments.

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(signif(size_diff$l[4]), 3), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(signif(size_diff$l[1]), 3), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by z-score for separated treatments."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], <!-- although this study extrapolated counts to full sample content, which is
rarely done in archaeological studies -->which could in part be due to the 
lack of oral amylase activity in our model.

@@

<!-- If you're reading this, thank you for taking the time to read the source 
documents. You're awesome! -->
We would like to thank Dr. Stephanie Schnorr for help with the amylase activity protocol.
We also thank everyone in the general vicinity
of the lab for enduring the smell of bacterial accumulation.
Since we did NOT make use of Sci-Hub to access articles stuck behind a paywall, 
we will NOT acknowledge the use of Sci-Hub in this study. <!-- Oops, did I just acknowledge it twice..? -->
